\input{preamble}

\title{On ARGs, pedigrees, and genetic relatedness matrices} % matri-cees

\author[1]{Brieuc Lehmann}
\author[2]{Hanbin Lee}
\author[3]{Luke Anderson-Trocm\'{e}}
\author[4]{Jerome Kelleher}
\author[5]{Gregor Gorjanc}
\author[6,7]{Peter L. Ralph}

\affil[1]{\small{Department of Statistical Science, University College London, WC1E 7HB, UK}}
\affil[2]{\small{Department of Statistics, University of Michigan, Ann Arbor MI 48109, USA}}
\affil[3]{\small{Department of Human Genetics, University of Chicago, Chicago IL 60637, USA}}
\affil[4]{\small{Big Data Institute, Li Ka Shing Centre for Health Information and Discovery, University of Oxford, OX3 7LF, UK}}
\affil[5]{\small{The Roslin Institute and Royal (Dick) School of Veterinary Studies, University of Edinburgh, UK}}
\affil[6]{\small{Institute of Ecology and Evolution, University of Oregon, Eugene OR 97402, USA}}
\affil[7]{\small{Department of Data Science, University of Oregon, Eugene OR 97402, USA}}

\date{\small{\today{}}}

\begin{document}

\maketitle

\section*{Abstract}

\input{abstract}

\section{Introduction}

\input{introduction}

% NOTE: the text for the "results" ended up in the methods 
% section because that was where most of the actual results
% were, and I'm trying to minimise the size of the diff to 
% avoid confusing Overleaf
\section{Results}

\input{methods}

\section{Methods}

\input{results}

\section{Discussion}

\input{discussion}

\section*{Data Availability}

\input{data_availability}

\section*{Acknowledgments}

\input{acknowledgements}

\bibliography{references}

\clearpage
% \section*{Appendix}

\appendix

\input{history}


\setcounter{table}{0}
\setcounter{figure}{0}
\renewcommand{\thetable}{A\arabic{table}}
\renewcommand{\thefigure}{A\arabic{figure}}

\section{Proof of correctness of Algorithm V}\label{sec-proof-algv-correct}

Here we prove that when Algorithm V completes, $v(s)$ is equal to the
$s^{\text{th}}$ entry of $\mathbf{C}\mathbf{w}$, as defined
in~\eqref{eq:similar_ralph2020}.
In fact, after each step in the algorithm (i.e., after each addition or removal of an edge), 
it is true that for \textit{every} node $n$, the sum of everything above that node is equal to the weighted sum of covariances
for that node including everything up to that point in the genome.
In other words, for every $n$,
\begin{align} \label{eqn:matvec_consistent}
    S_T(n):=
    \sum_{r \ge_T n} v(r) + z(r)w(r) 
    = 
    \sum_{h:h \le k} (b_h-b_{h-1}) \sum_{r:n \le_{T_h} r} \ell_{T_h}(r) \sum_{t: t \le_{T_h} r} w_t
\end{align}
where $T$ is the current tree.
This statement reduces to our claim that the algorithm is correct because
the final tree is an ``empty'' tree with no edges,
so at the end of the algorithm, the left-hand side is simply $v(n)+z(n)$.
This is again $v(n)$ because $z(n)$ is zero due to $x(n)=b_K$.
The right-hand side is equal to equation 
(\ref{eq:similar_ralph2020}) when $n$ is a sample node.

Each time we add an edge with child $c$ and parent $p$ to the tree (step V2), 
we add the value of $w(n)$ to $p$ and all nodes above $p$ in the tree;
when removing edges we subtract (step V1).
Since $w(n)$ is initialized so that each sample $s$ carries $w_s$, this ensures that $w(n)= \sum_{s \le n} w_s$ at all times [as in \cite{kelleher2016efficient, ralph2020efficiently}].

We prove that equation  (\ref{eqn:matvec_consistent}) is always true by induction.
At the first (empty) tree, this is certainly true,
as both sides are equal to zero.
We now consider \textbf{Step V3}. 
Tree $T$ and the bookkeeping variables $v$, $w$ and $x$ are left constant.
Advancing the position from $k$ to $k+1$ only changes 
$z(s)=\ell_T(s) (b_k - x(s))$ to 
$z'(s)=\ell_{T'}(s) (b_{k+1} - x(s))$.
Therefore, the appended value to the left-hand side is
\begin{align}
    \sum_{r: n \le_T r}z'(r)w(r) - z(r)w(r) = (b_{k+1}-b_{k})
    \sum_{r: n \le_T r} \ell_T(r) w(r)
\end{align}
which is also equal to the added value to the right-hand side.
Note that this is the only step that changes the value of equation (\ref{eqn:matvec_consistent}).

\textbf{Step V1} and \textbf{V2} leave the value of both sides unchanged.
Without the loss of generality, we prove this for \textbf{V2}.
Suppose that we added edge $(c,p)$ where $c$ and $p$ are the child and the parent nodes of the edge, respectively.
We can divide the nodes of $\nodes$ into four categories as (1) the child node $c$, (2) nodes below $c$,  (3) the parent node $p$ and nodes above $p$, and (4) all other nodes. 
The insertion changes the values of the intermediate arrays to $v'$, $w'$, $z'$, and $x'$ following \textbf{V2}.
We denote the new tree resulting from the insertion as $T'$.

Observe that the difference in the left-hand side after $\mathbf{V2}$ is
\begin{align}
    \begin{aligned}
        &\sum_{r:n \le_{T'} r} v'(r) + z'(r)w'(r) 
        - \sum_{r:n \le_{T} r} v(r) + z(r)w(r)
        \\
        &= \sum_{r: n\le_{T'}r, n \nleq_T r } v'(r) + z'(r)w'(r) \\
        &+ \sum_{r: n \le_{T'} r, n\le_T r} v'(r)+z'(r)w'(r)-v(r)-z(r)w(r) \\
        &- \sum_{r: n \nleq_{T'} r, n\le_T r} v(r) + z(r)w(r) \\
        &=  \sum_{r: n\le_{T'}r, n \nleq_T r } v'(r) + z'(r)w'(r)\\
        &+ \sum_{r: n\le_T r} v'(r)+z'(r)w'(r)-v(r)-z(r)w(r)
    \end{aligned}
    \label{preserve_split}
\end{align}
for node $n$.
The last line follows from $\{r: n \le_T r \} \subset \{r: n\le_{T'} r\}$ because $T'$ has more edges than $T$.
Nodes in each category have distinct values for the former and the latter sum of this equation.

The set $\{r: n \le_{T'} r, n \nleq_T r\}$ of the first summation is
\begin{align}
    \begin{cases}
         \{r: p \le_{T'} r\} & \quad \text{$n$ in category (1) or (2)} \\
        \emptyset & \quad \text{$n$ in category (3) or (4)}
    \end{cases}
\end{align}
This is because $p$ and the nodes above $p$ are the nodes that were previously not ancestors in $T$, but became ancestors of $c$ and those below $c$ after the addition of the new edge.
The set is empty for the nodes in the third and the fourth category because their ancestor nodes are unchanged after edge insertion.
Therefore, the first summation is
\begin{align}
    \begin{cases}
        \sum_{r: p \le_{T'} r} v'(r) + z'(r)w'(r) & \quad \text{$n$ in category (1) or (2)} \\
        0 & \quad \text{$n$ in category (3) or (4)}
    \end{cases}
\end{align}

The summand of the second summation $v'(r) + z'(r) - v(r) - z(r)$ is
\begin{align}
    \begin{cases}
    v'(c) - v(c) & \quad \text{$r$ in category (1)} \\
    0 & \quad  \text{$r$ in category (2), (3), or (4)}
    \end{cases} 
\end{align}
For $r=c$ ($r$ is in the first category), it follows from $z(c)=0$ due to $\ell_T(c)=0$ and $z'(c)=0$ due to $x'(c)=b_k$.
All the bookkeeping values of the second and the fourth category is unchanged by \textbf{V2}, so the summand is trivially zero.
When $r=p$ ($r$ belongs to the third category), $v'(r) = v(r) + z(r)w(r)$  and $z'(r)=0$ by the operations $v(r) \pluseq z(r)w(r)$ and $x(r)=b_k$ by \textbf{V2}.
Hence, the second summation is
\begin{align}
    \begin{cases}
        v'(c) - v(c) & \quad \text{$n$ in category (1) or (2)} \\
        0 & \quad \text{$n$ in category (3) or (4)}
    \end{cases}
\end{align}
because the set $\{r: n \le_T r\}$ contains $c$ if and only if $n$ belongs to either the third or the fourth category.
An expression for $v'(c)-v(c)$ comes from the operation $v(c) \minuseq v(r)$ following $v(r) \pluseq z(r)w(r)$ of \textbf{V2}
\begin{align}
    v'(c) - v(c) = - \sum_{r: p \le_{T} r} v(r) + z(r)w(r)
\end{align}

Combining the aforementioned results, we see that equation (\ref{preserve_split}) is 
\begin{align}
    \begin{cases}
        \sum_{r: p \le_{T'} r} v'(r) + z'(r)w'(r) - \sum_{r: p \le_{T} r} v(r) + z(r)w(r) & \text{$n$ $\in$ (1) or (2)} \\
        0 & \text{$n$ $\in$ (3) or (4)}
    \end{cases} 
\end{align}
Both cases reduces to zero because the set of nodes ancestral to $p$ are the same in $T$ and $T'$ ($\{r: p \le_T r\}=\{r: p \le_{T'} r\}$), 
 and $v'(r) = v(r) + z(r)w(r)$ for these nodes due to \textbf{V2} and $z'(r)=0$. 

The right-hand side remains the same because the operation changes nothing in its expression:
it's the working sum until the previous local tree that it does not contain any of 
the components of the current tree that is being modified.
This completes the proof.

% It is unclear what $v(\cdot)$ means intuitively in this version of the algorithm.
% It is merely an array designed for dynamic programming
% that carries intermediate numbers for each node.
% The way in which the algorithm modifies $v(\cdot)$ ensures that whenever we modify trees by deleting or inserting an edge, 
% the sum over $v(r)+z(r)w(r)$  for $ r \ge_{T}n $ equals to the working sum of the right-hand side of equation (\ref{eqn:matvec_consistent}).


\setcounter{table}{0}
\renewcommand{\thetable}{S\arabic{table}}
\setcounter{figure}{0}
\renewcommand{\thefigure}{S\arabic{figure}}
\setcounter{section}{1}
\renewcommand \thesection{S\arabic{section}}

\section*{Supplementary material}

\input{supplementary}

\end{document}
